---
title: "Vignette of the `inTextSummaryTable` package"
author: "Laure Cougnaud"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float:
      collapsed: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{inTextSummaryTable package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction

```{r options, echo = FALSE}
	
	library(knitr)
	opts_chunk$set(
		echo = TRUE, results = 'asis', warning = FALSE, 
		error = FALSE, message = FALSE, cache = FALSE,
		fig.width = 8, fig.height = 7,
		fig.path = "./figures_vignette/",
		#out.width = '0.7\\textwidth', 
		fig.align = 'center')#, out.height = 0.5\\textwidth', fig.path = 'graphs/') 
	options(width = 170)#, stringsAsFactors = FALSE
	options(warn = 1)#instead of warn = 0 by default -> to have place where warnings occur in the call to Sweave function
	
```

The package `inTextSummaryTable` contains functionalities to **create table of
summary statistics or counts of metrics of interest**, to **format them as
in-text table**, and to **export them to a`docx` document**.

```{r loadPackages}

	library(pander)
	library(inTextSummaryTable)

```

## Data format

The input data for the creation of summary table should be a **data.frame**,
usually loaded from _SAS_ data file (`sas7bdat` format).
The label of the variables stored in the `SAS` datasets is also used
for the title/captions. 

Example datasets from the _Pelican_ study included in the
`inTextSummaryTable` package itself are used in this vignette.  

Note that the `loadDataADaMSDTM` function of the `glpgUtilityFct` package can
be used to your custom `SAS` dataset(s) of interest.

```{r loadData}	
	
	# load example data
	data(ADaMDataPelicanInText)
	data(labelVarsADaMPelicanInText)
	
	dataAll <- ADaMDataPelicanInText
	labelVars <- labelVarsADaMPelicanInText

```

# Summary table

The main workhorse function is **`getSummaryStatisticsTable`** which creates a
in-text table of summary statistics for variable(s) of interest.

Two types of table can be created: a **count table for discrete variable**
(integer, factor, character) to indicate the counts of the number of subjects or
records per category of interest or **a summary statistics table to indicate
standard distribution statistics (mean, median, minimum, maximum, standard
error, standard deviation, and count statistics) for a numeric variable**.

By default, the type of the table is **automatically detected** based on the
variable type. This can also be specified via the `type` parameter set to:
`summaryTable`, `countTable` or `auto` (by default).

Note: all functionalities described for the summary statistics table are also
available for the count table (and vice versa).

## Summary statistics

The functionalities of the `getSummaryStatisticsTable` is demonstrated for the
summary statistics table.

### General

The _Pharmacokinetics_ data (`ADPP` dataset) is used as example for the summary
statistics table.

```{r summaryTable-PP-data}

	dataPP <- dataAll$ADPP
	
	# ensure that order of elements is the one specified in the corresponding numeric variable
	library(glpgUtilityFct)
	dataPP <- convertVarToFactor(
		data = dataPP, 
		var = c("PARCAT1", "PARAM", "AVISIT", "TRTP"), 
		varNum = c("PARCAT1N", "PARAMN", "AVISITN", "TRTPN")
	)

```

The variable to summarize is specified via the `var`
parameter. For the summary statistics table, this variable should be numeric.
**By default, an
entire set of summary statistical metrics are computed**, see
`? computeSummaryStatisticsTable` for the list of metrics.

```{r summaryTable-PP-simple}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL"
	)

```

### Row variable(s)

Specific grouping variable(s) for the rows can be specified via the
**`rowVar`** parameter.

For example, the summary statistics are computed for each `r labelVars["PARAM"]`
(`PARAM` parameter).

```{r summaryTable-PP-rowVar}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		rowVar = "PARAM"
	)
	
```

Grouping for the rows can be specified via **multiple nested variables** to the
`rowVar` parameter, which should be sorted in hierarchical order.

The `rowPadBase` parameter can be used to change the padding between each row
level.

```{r summaryTable-PP-rowVarMultiple}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		rowVar = c("PARCAT1", "PARAM"),
		rowPadBase = 10
	)
	
```

### Column variable(s)

Similarly, column(s) of interest are specified via the **`colVar`** parameter.
The total is included for each column in the header.

```{r summaryTable-PP-colVar}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = "TRTP"
	)
	
```

```{r summaryTable-PP-colVarMultiple}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = c("TRTP", "AVISIT")
	)
	
```

### Row and column combinations

```{r summaryTable-PP-rowColVar}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = c("TRTP", "AVISIT"),
		rowVar = c("PARCAT1", "PARAM")
	)
	
```

### Variable labels

The **labels used for the variables parameter** (row variables) **are
automatically extracted from the labels** contained in the _SAS_ dataset, by
specifying the `labelVars` parameter.

```{r summaryTable-PP-rowVarWithLabel-labelVars}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = c("TRTP", "AVISIT"),
		rowVar = c("PARCAT1", "PARAM"),
		labelVars = labelVars
	)
	
```

The label can also be specified directly via the `rowVarLab` parameter.

```{r summaryTable-PP-rowVarWithLabel-rowVarLab}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = c("TRTP", "AVISIT"),
		rowVar = c("PARCAT1", "PARAM"),
		rowVarLab = c('PARAM' = "Pharmacokinetics parameter", 'PARCAT1' = "Compound")
	)
	
```

### Custom statistics

For each element of the row and column variables, a set of summary statistics is
computed, see documentation of the `stats` parameter for the
`getSummaryStatisticsTable` function for more details.

**These statistics can be combined and the format can be customized via the
`stats` parameter** (`expression` or `call` object). 

In the case
that an unique statistic is specified, the 'Statistic' column doesn't appear in
the table.

For example, the summary table is restricted to the mean+-SE statistics.

```{r summaryTable-PP-meanSE}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = c("TRTP", "AVISIT"),
		rowVar = c("PARCAT1", "PARAM"),
		labelVars = labelVars,
		stats = list(expression(paste(formatC(Mean), "+-", formatC(SE)))),
		footer = "Statistics: mean +- SE",
		rowPadBase = 10
	)

```
The same table is created with the median (minimum, maximum) of the actual value
of .

```{r summaryTable-PP-medianMinMax}
	
	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = c("TRTP", "AVISIT"),
		rowVar = c("PARCAT1", "PARAM"),
		labelVars = labelVars,
		stats =  list(expression(paste0(formatC(Median), " (", formatC(Min), ", ", formatC(Max), ")"))),
		footer = "Statistics: median (min, max)"
	)

```

### Row as separated columns

**Row variables** that shouldn't be included as grouping for the rows, but as
**separated column** are via the `rowVarInSepCol` parameter.

```{r summaryTable-PP-rowVarInSepCol}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = c("TRTP", "AVISIT"),
		rowVar = c("PARCAT1", "PARAM"),
		rowVarInSepCol = "PARAM",
		labelVars = labelVars,
		stats =  list(expression(paste0(formatC(Median), " (", formatC(Min), ", ", formatC(Min), ")"))),
		footer = "Statistics: median (min, max)"
	)
	
```

### Summary statistics across columns

The **summary statistics across categories of the column variable**
are included via the parameter `colTotalInclude` set to TRUE.

```{r summaryTable-PP-colTotal}

	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = "TRTP",
		rowVar = c("PARCAT1", "PARAM"),
		labelVars = labelVars,
		colTotalInclude = TRUE,
		stats = list(expression(paste(formatC(Mean), "+-", formatC(SE)))),
		footer = "Statistics: mean +- SE",
		rowPadBase = 10
	)

```

### Multiple variables

Multiple variables to be summarized can be specified in the `var` parameter.

```{r summaryTable-multipleVariables}

	getSummaryStatisticsTable(
		data = subset(dataAll$ADLB, FASFL == "Y" & PARCAT1 == "CHEMISTRY"),
		var = c("AVAL", "CHG"),
		rowVar = c("PARCAT2", "PARAM"),
		colVar = c("TRTP", "AVISIT"),
		stats = list(expression(paste(formatC(Mean), "+-", formatC(SE)))),
		rowPadBase = 10,
		labelVars = labelVars,
		title = "Analysis value and change from baseline of the Chemical parameters"
	)

```

## Count table

Count table are obtained by specifying the **`type` parameter to 'countTable'**.
If **no variable is specified** (`var` parameter), the number of
subjects/records are counted in the **entire dataset**.
If a **variable is specified**, the **counts of each sub-category/class** of the
specified variable are extracted.

The adverse event dataset is used as an example.

```{r countTable-AE-data}

	dataAE <-  subset(dataAll$ADAE, FASFL == "Y")
	
	# ensure that order of elements is the one specified in the corresponding numeric variable
	library(glpgUtilityFct)
	dataAE <- convertVarToFactor(
		data = dataAE, 
		var = "TRTP", 
		varNum = "TRTPN"
	)

```

### General

By default, a set of custom statistics are computed, see the
`? computeSummaryStatisticsTable` for further details.

```{r countTable-AE-simple}

	getSummaryStatisticsTable(
		data = dataAE,
		type = "count"
	)

```	

### Extra functionalities

All functionalities provided for the summary statistics table are also available
for the count table.  

In the following example: the counts of subjects and records for the adverse
events are reformatted as: 'N/m (%)' with the `stats` parameter, and the adverse
event term are grouped according to the `r labelVars["AEHLGT"]` ('AEHLGT) and
`r labelVars["AEHLT"]` ('AEHLT') variable.

```{r countTable-AE-extra}

	getSummaryStatisticsTable(
		data = dataAE,
		type = "count",
		rowVar = c("AEHLGT", "AEHLT", "AELLT"),
		colVar = "TRTP",
		labelVars = labelVars,
		rowPadBase = 10,
		title = "Table: Adverse Events",
		stats = list(
			'N (%)' = expression(paste0(N, " (", round(PercN, 1), ")")),
			'm (%)' = expression(paste0(m, " (", round(Percm, 1), ")"))
		),
		footer = "Statistics: number and percentage of subjects: N (%) and records: m (%)"
	)

```	

### External dataset for total counts

By default, the **total per column is extracted from the number of subjects
available in the data used for the count table**. If the total should be
extracted from a **different dataset**, e.g. if only a subset of the patients
are present in the data, it should be specified via the **`dataTotal`** variable
(used also for the computation of the percentage).

For example, the total number of patients per treatment is extracted from the 
subject-level (`ADSL`) dataset. 

```{r countTable-LB-customDenominator}

	# dataset used to extract the 'Total'
	dataTotal <- dataAll$ADSL
	# should contain columns specified in 'colVar'
	dataTotal$TRTP <- dataTotal$TRT01P 
	
	getSummaryStatisticsTable(
		data = dataAE,
		type = "count",
		rowVar = c("AEHLGT", "AEHLT", "AELLT"),
		colVar = "TRTP",
		labelVars = labelVars,
		rowPadBase = 10,
		title = "Table: Adverse Events",
		dataTotal = dataTotal,
		stats = list(expression(paste0(N, " (", round(PercN, 1), ")"))),
		footer = "Statistics: N (%)"
	)

```

### Total count of events across rows

#### Entire table

The **total number of events across the entire table per column** is included if
the parameter **`rowTotalInclude`** is set to TRUE.

The label for the total is customized via the `rowTotalLab` parameter.

```{r countTable-LB-rowTotalInclude}

	# dataset used to extract the 'Total'
	dataTotal <- dataAll$ADSL
	# should contain columns specified in 'colVar'
	dataTotal$TRTP <- dataTotal$TRT01P 
	
	getSummaryStatisticsTable(
		data = dataAE,
		type = "count",
		rowVar = c("AEHLGT", "AEHLT", "AELLT"),
		colVar = "TRTP",
		labelVars = labelVars,
		rowPadBase = 10,
		title = "Table: Adverse Events",
		dataTotal = dataTotal,
		rowTotalInclude = TRUE, rowTotalLab = "Any adverse events",
		stats = list(expression(paste0(N, " (", round(PercN, 1), ")"))),
		footer = "Statistics: N (%)"
	)

```

#### Per group of the row variable

The total for **each sub-group of the row variable(s)** is now included via the
parameter **`rowSubtotalInclude`**.

```{r countTable-LB-rowSubtotalInclude}
	
	getSummaryStatisticsTable(
		data = dataAE,
		type = "count",
		rowVar = c("AEHLGT", "AEHLT", "AELLT"),
		colVar = "TRTP",
		labelVars = labelVars,
		rowPadBase = 10,
		title = "Table: Adverse Events",
		dataTotal = dataTotal,
		rowSubtotalInclude = TRUE,
		stats = list(expression(paste0(N, " (", round(PercN, 1), ")"))),
		footer = "Statistics: N (%)"
	)

```

### Empty rows/columns

The parameters **`rowInclude0` and `colInclude0`**
are set to TRUE to **include statistics for empty categories**.
These categories are specified via additional **levels in a factor variable**.

For example, the table of adverse events for white blood
cell analysis is created with an additional dummy treatment category (in
column) and extra analysis (in row)
category.

```{r countTable-LBWhiteBlood-emptyVars}

	## only consider White blood cell adverse events
	dataAEWBCA <- subset(dataAE, AEHLT == "White blood cell analyses")
	
	## create dummy categories for:
	# treatment
	dataAEWBCA$TRTP <- with(dataAEWBCA, 
		factor(TRTP, levels = c(unique(as.character(TRTP)), "Treatment B"))
	)
	# low-level term category
	dataAEWBCA$AELLT <- with(dataAEWBCA, 
		factor(AELLT, levels = c(unique(as.character(AELLT)), "Lymphocyte percentage increased"))
	)
	
	# create summary statistics table
	getSummaryStatisticsTable(
		data = dataAEWBCA,
		type = "count",
		rowVar = "AELLT",
		rowInclude0 = TRUE, colInclude0 = TRUE,
		colVar = "TRTP",
		labelVars = labelVars,
		rowPadBase = 10,
		title = "Table: Adverse Events: white blood cell analyses",
		dataTotal = dataTotal,
		stats = list(expression(paste0(N, " (", round(PercN, 1), ")"))),
		footer = "Statistics: N (%)"
	)

```

### Table filtering

#### Based on row variable

Only the AE categories occurring in at least 2 patients in at least one of the
low-level AR category are retained.

```{r countTable-AE-filterFct-rowVar1}

	library(plyr)
	filterFct <- function(x){
		ddply(x, "AELLT", function(y)
			if(any(y$N >= 2)) y
		)
	}
		
	getSummaryStatisticsTable(
		data = dataAE,
		type = "count",
		rowVar = c("AEHLGT", "AEHLT", "AELLT"),
		colVar = "TRTP",
		labelVars = labelVars,
		filterFct = filterFct,
		rowPadBase = 10,
		title = "Table: Adverse Events",
		stats = expression(paste0(N, " (", round(PercN, 1), ")")),
		footer = "Statistics: number and percentage of subjects: N (%)"
	)

```

#### Based on row and column variable

Only the AE categories occurring in at least 2 treated-patients in at least one
of the low-level AR category are retained.

```{r countTable-AE-filterFct-rowColVar}

	filterFct <- function(x){
		ddply(x, "AELLT", function(y)
			if(any(subset(y, TRTP == "treatmentX")$N >= 2))	y
		)
	}
		
	getSummaryStatisticsTable(
		data = dataAE,
		type = "count",
		rowVar = c("AEHLGT", "AEHLT", "AELLT"),
		colVar = "TRTP",
		labelVars = labelVars,
		filterFct = filterFct,
		rowPadBase = 10,
		title = "Table: Adverse Events",
		stats = expression(paste0(N, " (", round(PercN, 1), ")")),
		footer = "Statistics: number and percentage of subjects: N (%)"
	)

```

### Row ordering

The **categories in the row variables can be ordered** based on the
**`rowOrder`** variable:

* `alphabetical`: categories are ordered **alphabetically**
* `auto`: order of the categories in the **input variable** (factor levels) 
are retained (ordered alphabetically if not specifed)
* `total`: categories are ordered based on the **total of
  the sub-category**
  
#### Common order for all row variables

```{r countTable-LB-rowOrder-common}
	
	getSummaryStatisticsTable(
		data = dataAE,
		type = "count",
		rowVar = c("AEHLGT", "AEHLT", "AELLT"),
		rowOrder = "total",
		colVar = "TRTP",
		labelVars = labelVars,
		rowPadBase = 10,
		title = "Table: Adverse Events ordered based on total counts",
		dataTotal = dataTotal,
		rowSubtotalInclude = TRUE,
		stats = list(expression(paste0(N, " (", round(PercN, 1), ")"))),
		footer = "Statistics: N (%)"
	)

```

#### Different order for each row variables

In case the order should be different for each row variable, a named list is
provided for the `rowVar` parameter.

```{r countTable-LB-rowOrder-specific}
	
	dataAERowOrderByVar <- dataAE
	
	# order AE high-level term based on the code
	dataAERowOrderByVar <- convertVarToFactor(
		data = dataAERowOrderByVar, 
		var = "AEHLGT", varNum = "AEHLGTCD"
	)

	getSummaryStatisticsTable(
		data = dataAERowOrderByVar,
		type = "count",
		rowVar = c("AEHLGT", "AEHLT", "AELLT"),
		rowOrder = list(AEHLGT = "auto", 'AEHLT' = "total", AELLT = "alphabetical"),
		colVar = "TRTP",
		labelVars = labelVars,
		rowPadBase = 10,
		title = "Table: Adverse Events ordered based on total counts of high-level category",
		dataTotal = dataTotal,
		rowSubtotalInclude = TRUE,
		stats = list(expression(paste0(N, " (", round(PercN, 1), ")"))),
		footer = "Statistics: N (%)"
	)

```

### Row order based on the total of a column category

If the row categories should be **ordered by total counts for a specific
category of the column variable(s)**, a function **`rowOrderTotalFilterFct`** is
specified.

The categories of the adverse events of upper respiratory tract infections  of
are sorted based on their occurrence on only the treated
patients.

```{r countTable-LB-rowOrder-filterFct}

	dataAERowOrderFilterFct <- subset(dataAE, AEHLT == "Upper respiratory tract infections")
	getSummaryStatisticsTable(
		data = dataAERowOrderFilterFct,
		type = "count",
		rowVar = "AELLT",
		rowOrder = "total",
		colVar = "TRTP",
		labelVars = labelVars,
		rowPadBase = 10,
		rowOrderTotalFilterFct = function(x) subset(x, TRTP == "treatmentX"),
		title = "Table: Adverse Events ordered based on total counts",
		dataTotal = dataTotal,
		stats = list(expression(paste0(N, " (", round(PercN, 1), ")"))),
		footer = "Statistics: N (%)"
	)

```

## Mixed table

Summary statistics can be included for a **mixed table**, i.e. containing
**both numeric and discrete variables**.

The variable type is automatically detected by the function.

```{r mixedTable-SL}

	customStatsCat <- expression(paste0(N, " (", round(PercN, 1), ")"))
	customStatsNum <- expression(formatC(Mean), "+-", formatC(SE))
	
	getSummaryStatisticsTable(
		data = dataAll$ADSL,
		colVar = "TRT01P", 
		stats = list(
			RACE = customStatsCat, AGE = customStatsNum, SEX = customStatsCat,
			WEIGHTBL = customStatsNum, HEIGHT = customStatsNum,
			BMIBLGR1 = customStatsCat, SMOKE = customStatsCat
		),
		var = c("RACE", "AGE", "SEX", "WEIGHTBL", "HEIGHT", "BMIBLGR1", "SMOKE"),
		varLabGeneral = "Characteristic", varLabSubgroup = "Class",
		labelVars = labelVars,
		rowPadBase = 5,
		title = "Demographic table",
	)

```

# Use cases

## Laboratory abnormalities

The laboratory parameters are grouped by category, and the visit is indicated in
the rows: `PARCAT1`, `PARCAT2` and `AVISIT` are specified in `rowVar`. The
treatment is reported in the column: `TRTP` specified in `colVar`.

Because the laboratory measurement reported as normal or empty ('') in the ADaM
are not of interest, these classes are filtered via the `varIgnore` parameter.

By default, the number ('N') and percentage ('Perc') are reported for each
`rowVar`/`colVar`.

```{r example-countTable-LB}

	dataLB <- subset(dataAll$ADLB, AN01FL == "Y")
	
	# ensure that order of elements is the one specified in the corresponding numeric variable
	library(glpgUtilityFct)
	dataLB <- convertVarToFactor(
		data = dataLB, 
		var = c("PARCAT1", "PARCAT2", "AVISIT", "TRTP", "ANRIND"), 
		varNum = c("PARCAT1N", "PARCAT2N", "AVISITN", "TRTPN", "ANRINDN")
	)
	
	# only consider abnormalities
	dataLB <- subset(dataLB, !ANRIND %in% c("", "L"))
	
	# specific analysis visits
	dataLB <- subset(dataLB, AVISIT %in% c("Baseline", "Day 14", "Day 28", "Follow-up"))
	
	# create summary table
	summaryTableLBCount <- getSummaryStatisticsTable(
		data = dataLB,
		rowVar = c("PARCAT1", "PARCAT2", "PARAMCD", "ANRIND"),
		rowVarInSepCol = "ANRIND",
		colVar = c("TRTP", "AVISIT"),  
		type = "count",
		labelVars = labelVars,
		rowPadBase = 10,
		stats = list(expression(paste0(N, " (", round(PercN, 1), ")"))),
		title = "Table: Treatment-emergent laboratory abnormalities",
		file = "summaryStatisticsTable_LB_count.docx",
		footer = "Statistics: N (%)",
		landscape = TRUE
	)
	summaryTableLBCount

```

## Adverse events

```{r example-countTable-AE}

	dataAE <-  subset(dataAll$ADAE, FASFL == "Y")
	
	# ensure that order of elements is the one specified in the corresponding numeric variable
	library(glpgUtilityFct)
	dataAE <- convertVarToFactor(
		data = dataAE, 
		var = "TRTP", 
		varNum = "TRTPN"
	)

	# dataset used to extract the 'Total'
	dataTotal <- dataAll$ADSL
	# should contain columns specified in 'colVar'
	dataTotal$TRTP <- dataTotal$TRT01P 
	
	getSummaryStatisticsTable(
		data = dataAE,
		type = "count",
		rowVar = c("AEHLGT", "AEHLT", "AELLT"),
		colVar = "TRTP",
		labelVars = labelVars,
		rowPadBase = 10,
		title = "Table: Adverse Events",
		dataTotal = dataTotal,
		rowTotalInclude = TRUE,
		stats = list(expression(paste0(N, " (", round(PercN, 1), ")"))),
		footer = "Statistics: N (%)",
		file = "summaryStatisticsTable_AE_count.docx",
	)

```

## Pharmacokinetic parameters

```{r example-PP}
	
	dataPP <- dataAll$ADPP
	
	# ensure that order of elements is the one specified in the corresponding numeric variable
	library(glpgUtilityFct)
	dataPP <- convertVarToFactor(
		data = dataPP, 
		var = c("PARCAT1", "PARAM", "AVISIT", "TRTP"), 
		varNum = c("PARCAT1N", "PARAMN", "AVISITN", "TRTPN")
	)
	
	getSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		colVar = c("TRTP", "AVISIT"),
		rowVar = c("PARCAT1", "PARAM"),
		labelVars = labelVars,
		stats =  list(expression(paste0(formatC(Median), "\n(", formatC(Min), ", ", formatC(Min), ")"))),
		footer = "Statistics: median (min, max)",
		file = "summaryStatisticsTable_PP_summaryStatistics.docx"
	)

```


# Visualization

Summary statistics can be visualized via **error bars** by:

* extracting the summary statistics table with the
  `computeSummaryStatisticsTable` function (used for `getSummaryStatistics` function)
* visualizing the summary statistics via the `subjectProfileSummaryPlot`
  function

```{r visualization, out.width = "100%", fig.height = 6, fig.width = 9}

	summaryTableDf <- computeSummaryStatisticsTable(
		data = dataPP,
		var = "AVAL",
		rowVar = "PARAM",
		colVar = c("TRTP", "AVISIT")
	)

	# create the plot
	subjectProfileSummaryPlot(
		data = subset(summaryTableDf, !isTotal),
		xVar = "AVISIT",
		colorVar = "TRTP",
		labelVars = labelVars,
		facetVar = "PARAM",
		useLinetype = TRUE
	)

```

# Format details

The `getSummaryStatisticsTable` function **returns a `flextable` object
(`? flextable`)**.

When printed into the R console, this object is displayed in the browser.
This object can be inserted within a rmarkdown document by printing in the
specified document chunk.

This can also be exported to a _docx_ file, in case the `file` parameter is
specified.

# Appendix

## Session information

```{r includeSessionInfo, echo = FALSE}

	pander(sessionInfo())

```
